\chapter{Replication of `An Advanced Guide To Trade Policy Analysis'}

# Motivation

Despite solid theory and remarkable empirical success, the gravity equation is often applied a-theoretically and without account for estimation challenges that may render estimates biased/inconsistent.

Much of gravity models found in the literature fall under this quote from Leonardo da Vinci: *"He who loves practice without theory is like the sailor who boards ship without
a rudder and compass and never knows where he may cast."*

Traditionally, trade theory and trade policy analysis are performed in a general equilibrium setting. However, doing competent general equilibrium analysis is not a trivial task.

Modelers must know general-equilibrium theory so that their models have a sound theoretical basis; they must know how to solve their models; they need to be able to program (or at least communicate with programmers); they have to know about data sources and all their associated problems; and they have to be conversant with relevant
literature, especially that on elasticities [@shoven1984applied].

The gravity model has established itself as the ‘workhorse’ model in international trade, and thousands of papers have used it to study the effects of various determinants of trade.

Traditional determinants of trade flows are: Geography, Preferential Trade Agreements, Tariffs, Export Subsidies,
Non-tariff Measures, World Trade Organization Membership, Common Currency and Currency Unions, Foreign Direct
Investment, Immigration, Cultural Ties, Language, Colonial Ties.

Exotic determinants of trade flows are: Institutional Quality, Foreign Aid, Trust, Reputation for People, Reputation for Products, Export Promotion, Taxes, Mega Sporting Events (Olympic Games and World Cup), Embargoes, Other Trade Sanctions, Conflict, Wars, Piracy, and Ice Cap Melting.

The gravity model can simultaneously accommodate multiple countries and multiple sectors (and even firms). As such, the gravity framework can be used to capture the possibility that markets (sectors, countries, etc.) are linked and that trade policy changes in one market will trigger ripple effects in the rest of the world.

Structural Gravity translates the effects of the universe of bilateral trade policies to the country level and decomposes their incidence on producers and consumers in the world. Thus, the trade gravity model can be conveniently integrated within a wide class of broader general equilibrium models to study the links between trade, labor markets, investment, the environment, etc., while preserving tractability.

# Gravity models review

## Simple gravity model

The main reference for this section is @Woelver2018 and the references therein. Gravity models in their traditional form are inspired by Newton law of gravitation
\begin{equation*}
F_{ij}=G\frac{M_{i}M_{j}}{D^{2}_{ij}}.
\end{equation*}

The force $F$ between two bodies $i$ and $j$ with $i \neq j$ is proportional to the masses $M$ of these bodies and inversely proportional to the square of their geographical distance $D$. $G$ is a constant and as such of no major concern. The underlying idea of a traditional gravity model, shown for international trade, is equally simple
\begin{equation*}
X_{ij}=G\frac{Y_{i}^{\beta_{1}}Y_{j}^{\beta_{2}}}{D_{ij}^{\beta_{3}}}.
\end{equation*}

The trade flow $X$ is explained by $Y_{i}$ and $Y_{j}$ that are the masses of the exporting and importing country (e.g. the GDP) and $D_{ij}$ that is the distance between the countries. This is also used to study urban policies and migration flows.

Dummy variables such as common borders $contig$ or regional trade agreements $rta$ can be added to the model. Let $t_{ij}$ be the transaction cost defined as $t_{ij}= D_{ij} \exp(CONTIG_{ij} + RTA_{ij})$, therefore the model with frictions can be expressed as a log-linear model to use a standard estimation methods such as OLS
\begin{align*}
X_{ij} =& G\frac{Y_{i}^{\beta_{1}}Y_{j}^{\beta_{2}}}{t_{ij}^{\beta_{3}}} \\
\implies \log X_{ij} =& \beta_{0}\log G +\beta_{1}\log Y_{i}+\beta_{2}\log Y_{j}+\beta_{3}\log D_{ij}\\
&+ \beta_{4}CONTIG_{ij}+\beta_{5}RTA_{ij}.
\end{align*}

## Trade barriers model

Basically the model proposes that the exports $X_{ij}$ from $i$ to $j$ are determined by the supply factors in $i$, $Y_{i}$, and the demand factors in $j$, $Y_{j}$, as well as the transaction costs $t_{ij}$.

Next to information on bilateral partners $i$ and $j$, information on the rest of the world is included in the gravity model in the equation $Y=\sum_{i} Y_{i}= \sum_{j} Y_{j}$ that represents the worldwide sum of incomes (e.g. the world's GDP).

A key assumption is to take a fixed value $\sigma > 1$ in order to account for the preference for a variation of goods (e.g. in this model goods can be replaced for other similar goods).

The Multilateral Resistance terms are included via the terms $P$, Inward Multilateral Resistance, and $\Pi$, Outward Multilateral Resistance. The Inward Multilateral Resistance $P_i$ is a function of the transaction costs of $i$ to all trade partners $j$. The Outward Multilateral Resistance $\Pi_{j}$ is a function of the transaction costs of $j$ to all trade partners $i$ and their demand. 
The Multilateral Resistance terms dependent on each other. Hence, the estimation of structural gravity models becomes complex.

The econometric literature proposes the Multilateral Resistance model defined by the equations
\begin{align*}
X_{ij} =& \frac{Y_{i}Y_{j}}{Y}\frac{t_{ij}^{1-\sigma}}{P_{j}^{1-\sigma}\Pi_{i}^{1-\sigma}}\\
P_{i}^{1-\sigma} \text{ }& \sum_{j}\frac{t_{ij}^{1-\sigma}}{\Pi_{j}^{1-\sigma}}\frac{Y_{j}}{Y};\:\Pi_{j}^{1-\sigma}=\sum_{i}\frac{t_{ij}^{1-\sigma}}{P_{i}^{1-\sigma}}\frac{Y_{i}}{Y}
\end{align*}

## Model estimation

To estimate gravity equations you need a square dataset including bilateral flows defined by the argument `dependent_variable`, a distance measure defined by the argument `distance` that is the key regressor, and other potential influences (e.g. contiguity and common currency) given as a vector in `additional_regressors` are required.

Some estimation methods require ISO codes or similar of type character variables to compute particular country effects, and the rule of thumb for independent variables consists in that all dummy variables should be of type numeric (0/1) and f an independent variable is defined as a ratio, it should be logged.

# Models replication

## Work organization

The R package `tradepolicy` [@tradepolicy] allows easy summaries for partial and general equilibrium models. The detailed model replication is available from my ebook [Solutions Manual for An Advanced Guide to Trade Policy Analysis](https://r.tiid.org/R_structural_gravity/) published by UN ESCAP, and it's a material intended to provide a comprehensive explanation to reproduce the results from @yotov2016advanced in R. This ebook is not an attempt to give a thorough discussion of the theory behind gravity models, the references cited through the chapters shall fill those details.

From the considerations from Chapter 1, I had two audiences in mind: People who know R and are interested in learning about gravity models and people with no R knowledge who know gravity models theory. This assumes that the readers are familiar with linear regression, and that they shall read @yotov2016advanced and @wickham2016r alongside this material.

## Previous and side work

A previous work, `gravity` [@Woelver2018], intented for general gravity models estimation in R, provides a wrapper of different standard estimation methods for gravity models. This facilitates estimation of both log-log and multiplicative models.

In addition, `cepiigeodist` [@cepiigeodist] eases gravity modelling in R as it provides data on countries and their main city or agglomeration and the different distance measures and dummy variables indicating whether two countries are contiguous, share a common language or a colonial relationship. The reference article for these datasets is @CEPII201125.

## Partial equilibrium estimation

### Traditional gravity estimates

The most simple model considered in `tradepolicy` is an OLS estimation ignoring multilateral resistance terms
\begin{align*}
\log X_{ij,t} =& \beta_0 + \beta_1 DIST_{i,j} + \beta_2 CNTG_{i,j} + \beta_3 LANG_{i,j}\\ 
\text{ }& + \beta_4 CLNY_{i,j} + \beta_5 \log Y_{i,t} + \beta_6 \log E_{j,t} + \varepsilon_{ij,t}.
\end{align*}

The model can be summarised in the exact way as reported in @yotov2016advanced. The first challenge was to obtain country pair clustered standard errors for the estimated coefficients ($\hat{\beta}$), to provide an $R^2$ or pseudo-$R^2$ metric, alongside Ramsey Regression Equation Specification Error Test (RESET) test p-value. This mimics Stata regression output which is different from R results presentation. All of this is readily available in the `tp_summary_app1()` function.

```{r c2_ols_1, eval = FALSE}
tp_summary_app1(
  formula = "log_trade ~ log_dist + cntg + lang + clny + log_y + log_e",
  data = filter(ch1_application1, trade > 0),
  method = "lm"
)
```

```{r c2_ols_2, echo=FALSE, comment=NA}
load("~/tesis-magister/data/01-chapter1-traditional-gravity.RData")
ch1_app1_ols
```

In this output `etfe`/`itfe` refers to the presence of Exporter/Importer time fixed effects in the model and `nobs` is the number of rows used for the estimation after discarding unusable cases.

The package documentation covers all the details on each of the function arguments and the output, but here it is important to mention that clustered standard errors are not econometricians's sophistication. Estimations are clustered by trading pair in order to account for any intra-cluster correlations at the trading pair level. While important bilateral time-varying effects are controlled for with explanatory variables, such as RTAs, and any bilateral time-invariant effects are taken into account with fixed effects, some correlation pattern between pairs of countries over time may still be present in the error term. This correlation pattern is captured by clustering the errors over country-pairs @yotov2016advanced.

OLS estimation controlling with remote indexes and fixed effects are quite similar to this model, both don't pass the RESET test as the simple OLS model just shown. It is Pseudo-Poisson Maximum Likelihood (PPML) estimation the one that is interesting both statistically and economically. PPML proposes a multiplicative model that passes RESET test, and it requires to be estimated by using a Quasi-Poisson GLM.

PPML gravity model is defined as
\begin{align*}
X_{ij,t} =& \exp\left[\pi_{i,t} + \chi_{i,t} + \beta_1 \log(DIST)_{i,j} + \beta_2 CNTG_{i,j} +\right.\\
\text{ }& \left.\beta_3 LANG_{i,j} + \beta_4 CLNY_{i,j}\right] \times \varepsilon_{ij,t}.
\end{align*}

Where the added terms, with respect to the OLS model, are $\pi_{i,t}$ and $\chi_{i,t}$ that
account for exporter-time and importer-time fixed effects respectively.

The reason to compute this model even in spite of speed is that PPML is the only estimator that is perfectly consistent with the theoretical gravity model. By estimating with PPML the fixed effects correspond exactly to the corresponding theoretical terms.

And starting with this model, `tradepolicy` starts to show some advantages over fitting the models from scratch by using the `glm()` function. The pseudo-$R^2$ for this model needs to be computed after estimating the model because there is no $R^2$ for these models in R. The pseudo-$R^2$ is obtained as a function of the Kendall correlation between the observed and predicted values [@silva2006log].

The summary output, with hidden fixed effects, is 

```{r c2_ppml_1, eval = FALSE}
tp_summary_app1(
  formula = "trade ~ log_dist + cntg + lang + clny + exp_year + imp_year",
  data = ch1_application1,
  method = "glm"
)
```

```{r c2_ppml_2, echo=FALSE, comment=NA}
ch1_app1_ppml
rm(list = ls())
```

### The distance puzzle

The distance puzzle models accounts for a separation of the `log_dist` variable into multiple columns that account for discrete time effects. This is the model specification
\begin{align*}
X_{ij,t} =& \exp\left[\pi_{i,t} + \chi_{i,t} + \beta_1 \log(DIST)_{i,j} + \beta_2 CNTG_{i,j} + \beta_3 LANG_{i,j}\right]\times\\
\text{ }& \exp\left[\beta_4 CLNY_{i,j} + \beta_5 \log(DIST\_INTRA_{i,i})\right] \times \varepsilon_{ij,t}.
\end{align*}

The distance puzzle can be estimated via OLS or PPML. For OLS you need to exclude zero flows and loops in country pairs.

```{r ch2_dp_1, eval=FALSE}
tp_summary_app2(
  formula = "log_trade ~ 0 + log_dist_1986 + log_dist_1990 + log_dist_1994 +
    log_dist_1998 + log_dist_2002 + log_dist_2006 + cntg +
    lang + clny + exp_year + imp_year",
  data = filter(ch1_application2, importer != exporter, trade > 0),
  method = "lm"
)
```

```{r ch2_dp_2, echo=FALSE, comment=NA}
load("~/tesis-magister/data/02-chapter1-distance-puzzle.RData")
ch1_app2_ols
```

A useful metric for the decomposition of `log_dist` here is `pct_chg_log_dist`, which is the percentage change of the log distance coefficients as a time-varying effect. This is obtained by implementing the delta method on the distance coefficients weighted by the country pair clustered variance-covariance matrix.

Por PPML you only need to exclude loops, but keeping zerop flows. 
```{r, ch2_dp_3, eval=FALSE}
tp_summary_app2(
  formula = "trade ~ 0 + log_dist_1986 + log_dist_1990 +
    log_dist_1994 + log_dist_1998 + log_dist_2002 + log_dist_2006 +
    cntg + lang + clny + exp_year + imp_year",
  data = filter(ch1_application2, importer != exporter),
  method = "glm"
)
```

```{r ch2_dp_4, echo=FALSE, comment=NA}
ch1_app2_ppml
```

Other model variations, such as internal distance solution for the "distance puzzle", internal distance and home bias solution for the "distance puzzle" and the fixed effects solution for the "distance puzzle" are very similar to this approach.

### Regional trade agreements effects

This model specification includes gravity covariates, including both importer and exporter time fixed effects, and is a refinement over traditional gravity estimates model
\begin{align*}
X_{ij,t} =& \:\exp\left[\pi_{i,t} + \chi_{i,t} + \beta_1 \log(DIST)_{i,j} + \beta_2 CNTG_{i,j} + \beta_3 LANG_{i,j} +\right.\\
\text{ }& \:\left.\beta_4 CLNY_{i,j} + \beta_5 RTA_{ij,t}\right] \times \varepsilon_{ij,t}.
\end{align*}

The distance puzzle can be estimated via OLS or PPML, but based on the similarity with the previous models, you can go right into PPML model fitting starting with the most simplified case

```{r ch2_rta_1, eval=FALSE}
tp_summary_app3(
  formula = "trade ~ 0 + log_dist + cntg + lang + clny +
    rta + exp_year + imp_year",
  data = filter(ch1_application3, importer != exporter),
  method = "glm"
)
```

```{r ch2_rta_2, echo=FALSE, comment=NA}
load("~/tesis-magister/data/03-chapter1-regional-trade-agreements.RData")
ch1_app3_ppml
```

In this model summary we can see the same output as the previous model, with the addition of `total_rta_effect` which is just the sum of RTAs estimated coefficients and standard erro obtained with the Delta method. This model does not feature intra-national distance effects (`intr`).

Addressing for potential domestic trade diversion requires the inclusion of an international borders (`intl_brdr`) variable, a factor-type variable which starts to make the estimation slow and from here model estimation is computationally costly because of the number of estimated parameters, but this does not include general equilibrium effects yet.

```{r ch2_rta_3, eval=FALSE}
tp_summary_app3(
  formula = "trade ~ 0 + log_dist + cntg + lang + clny +
    rta + exp_year + imp_year + intl_brdr",
  data = ch1_application3,
  method = "glm"
)
```

```{r ch2_rta_4, echo=FALSE, comment=NA}
ch1_app3_intra
rm(list = ls())
```

These models constitute the basis for testing for potential "reverse causality" between trade and RTAs and addressing potential non-linear and phasing-in effects of RTAs.

## General equilibrium estimation

### Core CES model

The gravity model can simultaneously accommodate multiple countries and multiple sectors (and even firms). As such, the gravity framework can be used to capture the possibility that markets (sectors, countries, etc.) are linked and that trade policy changes in one market will trigger ripple effects in the rest of the world.

Structural Gravity translates the effects of the universe of bilateral trade policies to the country level and decomposes their incidence on producers and consumers in the world.

Some may argue that the great predictive power of the empirical gravity model is due to the use of a rich set of fixed effects $N \times T \times 2$, where $N$ denotes countries and $T$ denotes years. It's possible to replace the $N \times T \times 2$ dummies by two remoteness indexes to produce exactly the same results and fit [@wei1996intra, @baier2009estimating]. Thus, the predictive power of the empirical gravity model is indeed unprecedented.

The foundational stone of the structural gravity model is @anderson1979theoretical, which assumes that product differentiation exists by place of origin, that consumers globally have preferences that are truthfully represented by a Constant Elasticity of Substitution (CES) utility function, and that the equilibrium condition is a budget equilibrium. In this model, budgeting is defined in two stages, the lower level determines bilateral allocation and the upper level determines aggregate output and expenditures in each country.

Under Anderson's fundamental model, consumer's maximize their utility under budget constraints, which means to solve the next optimization problem
\begin{equation*}
\max \left\{ \sum_{i\in I} \beta_i^{\frac{1 - \sigma}{\sigma}} c_{ij}^{\frac{\sigma - 1}{\sigma}} \right\}^{\frac{\sigma - 1}{\sigma}} \quad \text{s.t.} \quad \sum_{i \in I} p_{ij} c_{ij} = E_j = Y_j.
\end{equation*}

Where

* $c_{ij}$ is consumption in destination $j$ of goods from source $i$
* $\sigma > 1$ is the elasticity of substitution across varieties/sources
* $\beta_i$ is a CES share parameter
* $p_{ij} = p_i t_{ij}$ is the price at destination $j$
* $p_i$ is the factory-gate/mill price at source $i$
* $t_ij > 1$ denotes trade frictions for shipping goods from $i$ to $j$
* $E_j$ is the total expenditure at destination $j$
* $Y_j = p_j Q_j$ is the total income at $j$, where $Q_j$ is the endowment

The target function can be maximized via Karush-Kuhn-Tucker algorithm with a restriction of the form $k - g(x) \geq 0$ for multiplier interpretation as shadow-price
\begin{equation*}
L = \left\{ \sum_{i\in I} \beta_i^{\frac{1 - \sigma}{\sigma}} c_{ij}^{\frac{\sigma - 1}{\sigma}} \right\}^{\frac{\sigma - 1}{\sigma}} + \mu \left( Y_j - \sum_{i \in I} p_{ij} c_{ij} \right).
\end{equation*}

Applying chain rule to obtain First Order Conditions (FOCs) with respect to $c_{ij}$
\begin{equation*}
\left\{ \sum_{i\in I} \beta_i^{\frac{1 - \sigma}{\sigma} - 1} c_{ij}^{\frac{\sigma - 1}{\sigma}} \right\}^{\frac{\sigma - 1}{\sigma}} \beta_i^{\frac{1 - \sigma}{\sigma} - 1} c_{ij}^{\frac{\sigma - 1}{\sigma} - 1} = \mu p_{ij}.
\end{equation*}

Repeating for consumption from $k$ , $c_{kj}$ , and then taking ratio of FOCs
\begin{equation*}
\frac{\beta_i}{\beta_k}^{\frac{1 - \sigma}{\sigma} - 1} \frac{c_{ij}}{c_{kj}}^{\frac{\sigma - 1}{\sigma} - 1} = \frac{p_{ij}}{p_{kj}}.
\end{equation*}

Solving for $c_kj$ in terms of $c_ij$, for any country $k$, and then replacing the solutions for all countries in the budget constraint and solving for $c_{ij}$, the solution to the consumer's problem is 
\begin{align*}
c_{ij} &= p_{ij}^{-\sigma} \left( \frac{\beta_i}{P_j} \right)^{1 - \sigma} E_j \\
P_j &= \left\{\sum_{i \in I} (\beta_i p_{ij})^{1 - \sigma} \right\}^{\frac{1}{1 - \sigma}}.
\end{align*}

Multiplying by $p_{ij}$ to get nominal demand $X_{ij}$ and defining $p_{ij} = p_i t_{ij}$
\begin{align*}
X_{ij} &= \left( \frac{\beta_i p_i t_{ij}}{P_j} \right)^{1 - \sigma} E_j \\
P_j &= \left\{\sum_{i \in I} (\beta_i p_{i} t_{ij})^{1 - \sigma} \right\}^{\frac{1}{1 - \sigma}}.
\end{align*}

Imposing market clearance condition 
\begin{equation*}
Y_i = \sum_{i \in I} X_{ij},\: \forall i \implies Y_i = \sum_{i \in I} \left( \frac{\beta_i p_i t_{ij}}{P_j} \right)^{1 - \sigma} E_j,\: \forall i
\end{equation*}

Dividing both sides by world income $Y$ and defining $\Pi$ as follows
\begin{align*}
\frac{Y_i}{Y} &= \sum_{i \in I} \left( \frac{\beta_i p_i t_{ij}}{P_j} \right)^{1 - \sigma} \frac{E_j}{Y},\: \forall i \\
\Pi^{1 - \sigma} &= \sum_{j \in J} \left(\frac{t_{ij}}{P_j}\right)^{1 - \sigma} \frac{E_j}{Y} \\
\implies \frac{Y_i}{Y} &= (\beta_i p_i \Pi_i)^{1 - \sigma},\: \forall i \\
\implies (\beta_i p_i)^{1 - \sigma} &= \frac{Y_i}{Y \Pi_i^{1 - \sigma}},\: \forall i
\end{align*}

This final term can be replaced in the expressions for $P_j$ and $X_{ij}$ to obtain the structural gravity system which can be estimated with GLMs even under market-clearing conditions
\begin{align*}
X_{ij} &= \frac{E_j Y_i}{Y} \left( \frac{t_{ij}}{P_j \Pi_i} \right)^{1 - \sigma} \\
P_j^{1 - \sigma} &= \sum_{i \in I} \frac{Y_i}{Y} \left( \frac{t_{ij}}{\Pi_i} \right)^{1 - \sigma} \\
\Pi_i^{1 - \sigma} &= \sum_{j \in I} \frac{E_j}{Y} \left( \frac{t_{ij}}{P_j} \right)^{1 - \sigma} \\
E_j &= Y_j = Q_j p_j \\
Y_j &= \sum_{i \in I} X_{ij}.
\end{align*}

In this model, $X_{ij}$ is the structural gravity equation, $P_j$ is the inward multilateral resistance, $\Pi_i$ is the outward multilateral resistance, $E_j$ is the expenditure/income/GDP and $Y_j$ defines the market clearance condition.

Most of newer articles build on this model. For example, @anderson2003gravity, another cornerstone article in gravity literature, builds up from this model.

For the purpose of estimation, the structural gravity system can be expressed as
\begin{align*}
X_{ij} &= \frac{E_j Y_i}{Y} \left( \frac{t_{ij}}{P_j \Pi_i} \right)^{1 - \sigma} \label{eq:1}\\
P_j^{1 - \sigma} &= \sum_{i \in I} \frac{Y_i}{Y} \left( \frac{t_{ij}}{\Pi_i} \right)^{1 - \sigma} \\
\Pi_i^{1 - \sigma} &= \sum_{j \in I} \frac{E_j}{Y} \left( \frac{t_{ij}}{P_j} \right)^{1 - \sigma} \\
E_j &= Y_j = Q_j p_j \\
p_j &= \frac{(Y_j / Y)^{\frac{1}{1 - \sigma}}}{\beta_j \Pi_j}.
\end{align*}

### General equilibrium PPML model

From the core model, it is possible to estimate a General Equilibrium PPML (GEPPML) model [@yotov2016advanced]. There are four sequential steps that define an 'auction' or 'russian doll' process.

**Step 1: Solve the Baseline Gravity Model**

Step 1A is to obtain estimates of trade costs and trade elasticities. This is done by using PPML, panel data with intervals, exporter-time fixed effects, importer-time fixed effects, and pair fixed effects to estimate
\begin{equation*}
X_{ij,t} = \exp[T_{ij,t}\beta + \pi_{i,t} + \xi_{j,t}] \times \varepsilon_{ij,t}.
\end{equation*}

Accounting for perfect collinearity, imposing a normalization for the MRs, and estimating structural gravity without a constant implies: $\pi_{i,t} = Y_{i,t} / (\Pi^{1-\sigma}_{i,t} Y_t)$ and $\xi_{0,t} = E_{0,t}$, where $\xi_{0,t} = E_{0,t}$ is the dropped fixed effect.

For this step it is crucial to select a country with good data, since it shall be the reference factor for the model. It is also important to select a country that is ‘remote’ from the counterfactual. The estimation can be done with any method from the literature, such as the tetrads method already implemented in R [@Woelver2018].

Step 1B is to construct baseline indexes. This is done by using the estimates of the gravity fixed effects together with data on output and expenditure to construct the multilateral resistances
\begin{align*}
\Pi^{1-\sigma}_{1,t} &= \frac{Y_{i,t}}{\exp(\hat{\pi}_{i,t})} \times \frac{E_{0,t}}{Y_t} \\
P^{1-\sigma}_{j,t} &= \frac{E_{j,t}}{\exp(\hat{\xi}_{j,t})} \times \frac{1}{E_{0,t}}.
\end{align*}

Where, by construction, $Y_{i,t} = \sum_j X_{ij,t}$ and $E_{j,t} = \sum_i X_{ij,t}$. Any other baseline GE index can be constructed, being it real consumption, terms of trade or others.

A starting point, based on partial equilibrium results, is to fit the model
\begin{align*}
X_{ij,t} =& \:\exp\left[\pi_{i,t} + \chi_{i,t} + \beta_1 \log(DIST)_{i,j} + \beta_2 CNTG_{i,j} + \beta_3 INTL_{i,j}\right] \times \varepsilon_{ij,t}.
\end{align*}

The code for this model is straightforward.

```{r ch2_app1_baseline_1, eval = FALSE}
fit_baseline_app1 <- glm(
  trade ~ 0 + log_dist + cntg + intl + exporter + importer,
  family = quasipoisson(link = "log"),
  data = ch2_application1
)
```

One challenge inR is to extract the estimated fixed effects coefficients in order to construct the variables for export and import fixed effects. This was done by implementing the `tp_fixed_effects()` function.

```{r ch2_app1_baseline_3, eval = FALSE}
ch2_application1 <- ch2_application1 %>%
  left_join(
    tp_fixed_effects(fit_baseline_app1),
    c("exporter", "importer")
  )
```

After performing the steps to obtain both OMRs and IMRS, to complete this stage of the estimation, I had to create a column with the estimated international trade for given output and expenditures. This starts by adding a column, `tradehat_bln`, with the regression output, and then grouping by exporter and summarising to obtain the required column `xi_bln`.

**Step 2: Define a counterfactual experiment**

Consists in defining the policy experiment by changing the definition of the policy variables in vector $T_{ij,t}. Alternatively, one may change the values for the trade cost elasticities, the $\beta$'s.

On the one hand, one alternative is to eliminate the border variable and then generate the logged trade costs used in the constraint.

```{r ch2_app1_counterfactual_1, eval = FALSE}
ch2_application1 <- ch2_application1 %>%
  mutate(
    tij_cfl = exp(fit_baseline_app1$coefficients["log_dist"] * log_dist +
                  fit_baseline_app1$coefficients["cntg"] * cntg),
    log_tij_cfl = log(tij_cfl)
  )
```

On the other, a second alternative is to define a new counterfactual border variable.

```{r ch2_app1_counterfactual_2, eval = FALSE}
ch2_application1 <- ch2_application1 %>%
  mutate(
    intl_cfl = 0,
    tij_bln = exp(fit_baseline_app1$coefficients["log_dist"] * log_dist +
                  fit_baseline_app1$coefficients["cntg"] * cntg +
                  fit_baseline_app1$coefficients["intl"] * intl_cfl),
    log_tij_cfl = log(tij_cfl)
  )
```

**Step 3: Solve the counterfactual model**

Step 3A is to obtain conditional GE effects
\begin{equation*}
X_{ij,t} = \exp[T_{ij,t}^{CFL}\bar{\beta} + \pi_{i,t}^{CFT} + \xi_{j,t}^{CFT}] \times \varepsilon_{ij,t}^{CFT}.
\end{equation*}

Where, the bar symbol is used to capture the fact that certain coefficients are constrained to be equal to their estimated counterparts from the equation in step 1 or to their new values as defined in step 2.

This step is completed by repeating step 1A with the new gravity fixed effects estimates and the original data on output and expenditure to construct the conditional GE estimates of the MRs and construct any other GE indexes of interest.

Step 3B is to estimate the full endowment gravity model. This step delivers estimates of the fixed effects that
correspond to the full endowment scenario. This step is performed in a loop with four stages: (i) Allow for changes in factory-gate prices; (ii) Allow for endogenous income, expenditure, and trade; (iii) Estimate the full endowment model; (iv) Construct full endowment GE Effects.

In practice, this steps is a recursion by fitting a model similar to the model from step 1, the constrained gravity model, where $\pi_{j,t}$ and $\chi_{j,t}$ are altered
\begin{align*}
X_{ij,t} =& \:\exp\left[\pi_{i,t}^{CFL} + \chi_{i,t}^{CFL} + \beta_1 \log(DIST)_{i,j} + \beta_2 CNTG_{i,j} + \beta_3 INTL_{i,j}\right] \times \varepsilon_{ij,t}.
\end{align*}

A large advantage from R is to have the `offset()` function, which makes this estimation of a counterfactual model very straightforward.

```{r ch2_app1_counterfactual_3, eval = FALSE}
fit_counterfactual_app1 <- glm(
  trade ~ 0 + exporter + importer + offset(log_tij_cfl),
  family = quasipoisson(link = "log"),
  data = ch2_application1
)
```

From this, it's direct to compute the conditional general equilibrium effects of trade, as this is a repetition of step 1. The problem is to construct the loop with an iterative procedure to converge to full endowment general equilibrium effects, this is the 'auction' process implicit in GE. 

Taking the elasticities of substitution from the literature, and building the price changes in a counterfactual scenario, the problem reduces to run the mentioned loop of $N$ steps, which cannot be divided into smaller pieces because the step $N$ depends on the step $N-1$.

```{r ch2_app1_counterfactual_11, eval = FALSE}
# set parameters
max_dif <- 1; sd_dif <- 1; change_price_i_old <- 0

i <- 1
while(sd_dif > 1e-5 | max_dif > 1e-5) {
  ch2_application1 <- ch2_application1 %>%
    mutate(trade_1 = tradehat_0 * change_p_i_0 * change_p_j_0 / 
             (change_omr_full_0 * change_imr_full_0))

  # repeat the counterfactual model +
  # compute the conditional general equilibrium effects of trade
  # compute the change in prices for exporters and importers +
  # compute both outward and inward multilateral resistance +
  # compute the conditional general equilibrium effects of trade
  ...
  
  # update the differences
  max_dif <- abs(max(ch2_application1$change_p_i_0 - change_price_i_old))
  sd_dif <- sd(ch2_application1$change_p_i_0 - change_price_i_old)
  change_price_i_old <- ch2_application1$change_p_i_0

  # compute changes in outward and inward multilateral resistance
  ch2_application1 <- ch2_application1 %>%
    mutate(
      change_omr_full_1 = omr_cfl_1 / omr_cfl_0,
      omr_cfl_0 = omr_cfl_1,
      ...
    )

  i <- i + 1
}
```

The rest of the step consists in computing the full endowment general equilibrium of trade. This short, but exhaustive summary, should have made clear that taking GLM's output as inputs for more GLM models is computationally costly

**Step 4: Collect, construct and report indexes of interest**

This step is direct. It consists in computing the change in full endowment general equilibrium factory-gate price on export side (`change_price_full`), the change in conditional and full general equilibrium outward multilateral resistances (`change_omr_*`), and the change in conditional and full general equilibrium international trade (`change_xi_*`).

In particular, the first GE model from @yotov2016advanced results, converges after 28 iterations, resulting in full replication of the book's figures and models output.

```{r ge_plot, echo=FALSE, fig.height=3.5, fig.width=5, message=FALSE, warning=FALSE, dpi=300}
load("~/tesis-magister/data/04-chapter2-trade-without-borders.RData")

ggplot(data = ch2_application1 %>%
         filter(exporter != "HKG")) +
  geom_point(aes(x = log_y, y = change_xi_cfl, color = "1")) +
  geom_point(aes(x = log_y, y = change_xi_full, color = "2")) +
  labs(
    x = "Log value of output",
    y = "Percent change of exports",
    title = "Effects of abolishing international borders on exports",
    caption = "Source: Authors' calculations",
    color = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(
    labels = c(
      "Conditional general equilibrium",
      "Full endowment general equilibrium"
    ),
    values = c("#b6b8dd","#232958")
  )
```

```{r ge_plot2, echo=FALSE, fig.height=3.5, fig.width=5, message=FALSE, warning=FALSE, dpi=300}
ggplot(data = ch2_application1 %>% filter(exporter != "HKG")) +
  geom_point(aes(x = log_y, y = change_imr_full, color = "1")) +
  geom_point(aes(x = log_y, y = change_price_full, color = "2")) +
  geom_point(aes(x = log_y, y = rgdp, color = "3")) +
  labs(
    x = "Log value of output",
    y = "Percent changes",
    title = "Effects of abolishing international borders on real GDP",
    caption = "Note: The inward multilateral resistances have been reformulated\nby multiplying their value by minus one.\nSource: Authors' calculations",
    color = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(
    labels = c(
      "-(inward multilateral resistances)",
      "Factory-gate price",
      "Real GDP"
    ),
    values = c("#3bade3", "#b6b8dd", "#232958")
  )

rm(list = ls())
```

## Findings

It was possible to fully replicate each model, figure and summary from @yotov2016advanced. The solely differences were model intercepts because of how Stata and R differ at some steps of GLMs fitting, but the slopes, standard errors and every model assessment metric could be fully replicated.

The replication work was somewhat straightforward because of the availability if: (i) code and datasets, both documented; (ii) comprehensive explanations of the theory, which allowed to provide a meaningful translation from Stata to R, even when the translation was not 1:1 because of difference on how both languages deal with categorical variables; (iii) direct communication with the principal author, who replied all of my questions about the data cleaning steps.
